Regression models for Cylindrical data in Psychology

### 2019-07-26 Manuscript/Manuscript.Rmd
Manuscript update
Corrections Christophe

author: 
  - name          : "Jolien Cremers"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : ""
    email         : "joliencremers@gmail.com"
  - name          : "Helena J.M. Pennings"
    affiliation   : "2,3"
  - name          : "Christophe Ley"
    affiliation   : "4"

affiliation:
  - id            : "1"
    institution   : "Department of Methodology and Statistics, Utrecht University"
  - id            : "2"
    institution   : "TNO"
  - id            : "3"
    institution   : "Department of Education, Utrecht University"
  - id            : "4"
    institution   : "Department of Applied Mathematics, Computer Science and Statistics, Ghent University"

@@

abstract: |
  Cylindrical data are multivariate data which consist of a directional,
  in this paper circular, and a linear component. Examples of cylindrical
  data in psychology include human navigation (direction and distance of 
  movement), eye-tracking research (direction and length of saccades) and
  data from an interpersonal circumplex (type and strength of
  interpersonal behavior). In this paper we adapt four models for
  cylindrical data to include a regression of the circular and linear
  component onto a set of covariates. Subsequently, we illustrate how to
  fit these models and interpret their results on a dataset on the
  interpersonal behavior of teachers.

@@

figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

@@

```{r, echo = FALSE}
library(papaja)
library(circular)
library(bpnreg)
library(haven)
library(tikzDevice)
library(plotrix)
library(tidyr)
library(dplyr) #sample_n()
library(MASS) #mvrnorm()
library(kableExtra) #latex tables
library(plotrix)
library(caret) #cross-validation
library(coda) #geweke diagnostic
```
\vspace{-0.75cm}
In social sciences the use of cylindrical data is very common. Such data consist
of a linear and a circular component. @gurtman2011reasoning refers to such data
as vectors, with a directional measure (i.e., the circular component) and a
measure indicating the magnitude (i.e., the linear component). Many established
models in psychology are often referred to as circular or circumplex models, but
those models are cylindrical. Examples of such cylindrical models are the
interpersonal circle/circumplex
[@leary1957;@wiggins1996history;@wubbels2006interpersonal], the circumplex of
affect [@russell1980circumplex], the circumplex of human emotion
[@plutchik1997general] or the model of human values [@schwartz1992values].
\newline
\indent Also, many of the more recent types of data that are studied in psychology are
cylindrical. For example, research on human navigation uses data where distance
(i.e., the linear component) and direction (i.e., the circular component) are of
interest [@chrastil2017rotational] or in eye-tracking, the saccade data also
consist of both the direction (i.e., the circular variable) and the duration
(i.e., the linear variable) (e.g., @rayner200935th). Apart from the social
sciences, data with a circular and linear outcome more commonly occur in
meteorology [@garcia2013exploring], ecology [@garcia2014test] or
marine research [@lagona2015hidden]
\newline
\indent Up until now researchers studying cylindrical data had to rely on linear
statistical methods to analyze their research results. However, lately more and more
of these researchers acknowledge that linear methods are not sufficient and call
for new methods [@gurtman2011reasoning; @pennings2017complexity;
@wright2009integrating] that take into account both the circular and the linear
component of these data.
\newline
\indent The aim in the present paper is twofold. Firstly, we intend to fill
the above mentioned gap in the literature by showing that the use of cylindrical
models can benefit the analysis of circumplex data and cylindrical data in
psychology in general. More specifically we will show these benefits for
interpersonal teacher data from the field of educational psychology. Apart from
modelling the dependence between the linear and circular component of a
cylindrical variable we would also like to predict the two components from a set
of covariates in a regression model. Our second aim therefore is to adapt
several existing cylindrical models in such a way that they include a regression
of both the linear and circular component of a cylindrical variable onto a set
of covariates. From now on we will therefore refer to the components of the
cylindrical variable as outcome components. These adapted cylindrical models are
then used  to analyse the teacher data.

\vspace{-0.75cm}
\subsection{Cylindrical Data}
\vspace{-0.75cm}
Data that consist of a linear variable and a circular variable is called
cylindrical data. A circular variable is different from the linear variable in
the sense that it is measured on a different scale. Figure \ref{circline} shows
the difference between a circular scale (right) and a linear scale (left). The
most important difference is that on a circular scale the datapoints 0\(^\circ\)
and 360\(^\circ\) are connected and in fact represent the same number while on a
linear scale the two ends, \(-\infty\) and \(\infty\), are not connected and
consequently the values 0\(^\circ\) and 360\(^\circ\) are located on different
places on the scale. Both circular data and cylindrical data require special
analysis methods due to this periodicity in the scale of a circular variable
(see e.g. @fisher1995statistical for an introduction to circular data and
@mardia2000directional, @jammalamadaka2001topics and @ley2017modern for a more
elaborate overview).\newline
\indent In the literature, several methods have been put forward to model the
relation between the linear and circular component of a cylindrical variable.
Some of these are based on regressing the linear component onto the circular
component using the following type of relation: \[y = \beta_0 +
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is the
linear component and \(\theta\) the circular component [@mardia1978model;
@johnson1978some; @mastrantonio2015bayesian]. Others model the relation in a
different way, e.g. by specifying a multivariate model for several linear and
circular variables and modelling their covariance matrix
[@mastrantonio2018joint] or by proposing a joint cylindrical distribution. For
example, @abe2017tractable introduce a cylindrical distribution based on a
Weibull distribution for the linear component and a sine-skewed von Mises
distribution for the circular component and link these through their respective
shape and concentration parameters. However, none of the methods that have been
proposed thus far include additional covariates onto which both the circular and
linear component are regressed.

@@

\vspace{-0.75cm}
\section{Teacher Data}\label{Example}
\vspace{-0.75cm}
\begin{figure}
\centering
\includegraphics[width = 0.8\textwidth]{Plots/IPC-T.png}
\caption{The interpersonal circle for teachers (IPC-T). The words presented in
the circumference of the circle are anchor words to describe the type of
behavior located in each part of the IPC.}
\label{QTI}
\end{figure}

The motivating example for this article comes from the field of educational
psychology and was collected for the studies on classroom climate of
@vanderWant2015role, @Claessens2016side and @pennings2017phd. An
indicator of the quality of the classroom climate is the students' perception of
their teachers' interpersonal behavior. These interpersonal perceptions, both in
educational psychology as well as in other areas of psychology, can be measured
using circumplex measurement instruments (see @horowitz2010handbook for an
overview of many such instruments).\newline
\indent The circumplex data used in this paper are measured using the
Questionnaire on Teacher Interaction (QTI) [@wubbels2006interpersonal] which is
one such circumplex measurement instrument. The QTI is designed to measure
student perceptions of their teachers' interpersonal behavior and contains items
that load on two interpersonal dimensions: Agency and Communion. Agency refers
to the degree of power or control a teacher exerts in interaction with his/her
students. Communion refers to the degree of friendliness or affiliation a
teacher conveys in interaction with his/her students. The loadings on the two
dimensions of the QTI can be placed in a two-dimensional space formed by Agency
(vertical) and Communion (horizontal), see Figure \ref{QTI}. This space is
called the interpersonal circle/circumplex (IPC) and different parts of this
space are characterized by different teacher behavior, e.g. 'helpful' or
'uncertain'. The IPC is ``a continuous order with no beginning or end''
[@gurtman2009exploring, p. 2]. We call such ordering a circumplex ordering and
the IPC is therefore often called the interpersonal circumplex. The ordering
also implies that scores on the IPC could be viewed as a circular variable. This
circular variable represents the type of interpersonal behavior that a teacher
shows towards his/her students.\newline

>>>>

author: 
  - name          : "Jolien Cremers"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : ""
    email         : "jcre@sund.ku.dk"
  - name          : "Helena J.M. Pennings"
    affiliation   : "2,3"
  - name          : "Christophe Ley"
    affiliation   : "4"

affiliation:
  - id            : "1"
    institution   : "Department of Biostatistics, University of Copenhagen"
  - id            : "2"
    institution   : "TNO"
  - id            : "3"
    institution   : "Department of Education, Utrecht University"
  - id            : "4"
    institution   : "Department of Applied Mathematics, Computer Science and Statistics, Ghent University"

@@

abstract: |
  Cylindrical data are multivariate data which consist of a directional,
  in this paper circular, and a linear component. Examples of cylindrical
  data in psychology include human navigation (direction and distance of 
  movement), eye-tracking research (direction and length of saccades) and
  data from an interpersonal circumplex (location and intensity on the IPC). 
  In this paper we adapt four models for
  cylindrical data to include a regression of the circular and linear
  component onto a set of covariates. Subsequently, we illustrate how to
  fit these models and interpret their results on a dataset on the
  interpersonal behavior of teachers.

@@

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes
draft             : no

@@

```{r, echo = FALSE}
library(papaja)
library(circular)
library(bpnreg)
library(haven)
library(tikzDevice)
library(plotrix)
library(tidyr)
library(dplyr) #sample_n()
library(MASS) #mvrnorm()
library(kableExtra) #latex tables
library(plotrix)
library(caret) #cross-validation
library(coda) #geweke diagnostic
```
In the social sciences the use of cylindrical data is very common. Such data
consist of a linear and a circular component. @gurtman2011reasoning refers to
such data as vectors, with a directional measure (i.e., the circular component)
and a measure indicating the magnitude (i.e., the linear component). Many
established models in psychology are often referred to as circular or circumplex
models, but those models are cylindrical. Examples of such cylindrical models
are the interpersonal circle/circumplex
[@leary1957;@wiggins1996history;@wubbels2006interpersonal], the circumplex of
affect [@russell1980circumplex], the circumplex of human emotion
[@plutchik1997general] or the model of human values [@schwartz1992values].
\newline
\indent Also, many of the more recent types of data that are studied in
psychology are cylindrical. For example, research on human navigation uses data
where distance (i.e., the linear component) and direction (i.e., the circular
component) are of interest [@chrastil2017rotational] or in eye-tracking, the
saccade data also consist of both the direction (i.e., the circular variable)
and the duration (i.e., the linear variable) (e.g., @rayner200935th). Apart from
the social sciences, data with a circular and linear component more commonly
occur in meteorology [@garcia2013exploring], ecology [@garcia2014test] or marine
research [@lagona2015hidden] \newline
\indent Up until now researchers studying cylindrical data had to rely on linear
statistical methods to analyze their research results. However, lately more and
more of these researchers acknowledge that linear methods are not sufficient and
call for new methods [@gurtman2011reasoning; @pennings2017complexity;
@wright2009integrating] that take into account both the circular and the linear
component of these data.
\newline
\indent The aim in the present paper is twofold. Firstly, we intend to fill the
above mentioned gap in the literature by showing that the use of cylindrical
models can benefit the analysis of circumplex data and cylindrical data in
psychology in general. More specifically we will show these benefits for
interpersonal teacher data from the field of educational psychology. Apart from
modelling the dependence between the linear and circular component of a
cylindrical variable we would also like to predict the two components from a set
of covariates in a regression model. Our second aim therefore is to adapt
several existing cylindrical models in such a way that they include a regression
of both the linear and circular component of a cylindrical variable onto a set
of covariates. These adapted cylindrical models are then used  to analyse the
teacher data.

\section{Modelling Framework}
Data that consist of a linear variable and a circular variable are called
cylindrical data. The circular variable is different from the linear variable in
the sense that it is measured on a different scale. Figure \ref{circline} shows
the difference between a circular scale (right) and a linear scale (left). The
most important difference is that on a circular scale the datapoints 0\(^\circ\)
and 360\(^\circ\) are connected and in fact represent the same number while on a
linear scale the two ends, \(-\infty\) and \(\infty\), are not connected and
consequently the values 0\(^\circ\) and 360\(^\circ\) are located on different
places on the scale. Both circular data and cylindrical data require special
analysis methods due to this periodicity in the scale of a circular variable
(see e.g. @fisher1995statistical for an introduction to circular data and
@mardia2000directional, @jammalamadaka2001topics and @ley2017modern for a more
elaborate overview).

\hfil [Figure \ref{circline} about here] \hfil

\indent From now on we will refer to the circular random variable as $\Theta$,
the circular component, and the linear random variable as $Y$, the linear
component, with realizations $\theta_i$ and $y_i$ for each measurement $i=1,
\dots, n$ where $n$ is the sample size. The circular component is measured in
radians and consequently takes values in the interval $[0, 2\pi)$. Note that the
round brackets mean that $2\pi$ is not included in the interval since this
represents the same value as 0 as a result of periodicity. Radians can be
converted to degrees using the relation $1 \:\text{rad} =
1*\frac{180^\circ}{\pi}$. A circular variable is usually described using its
first two moments, circular location or mean $\mu$ and mean resultant length
$\rho$\footnote{In the von Mises distribution, a common distribution for a
circular outcome, we have a concentration parameter $\kappa$ that is related to
$\rho$ as $\rho = A_1(\kappa)$, where $A_1{\kappa} = I_1(\kappa)/I_0(\kappa)$
and $I_0()$ and $I_1()$ are modified Bessel functions of order 0 and 1,
respectively.}. The sample values of these two parameters are referred to as
$\bar{\theta}$ and $\hat{\rho}$. The mean resultant length lies between 0,
meaning the data is not concentrated at all \emph{i.e.} spread over the entire
circle, and 1, meaning all data is concentrated at a single point on the circle.
The linear component is measured on the real line and can take values on
$(-\infty, \infty)$ or we can constrain it to the positive real line $(0,
\infty)$. Depending on which distributional assumptions are used we describe the
distribution of the linear component using a mean $\mu$ and variance $\sigma^2$
or a scale $\nu$ and shape $\alpha$ parameter. When skewed distributions are
used for either the circular or linear component we may also include a skewness
parameter $\lambda$.\newline
\indent In a cylindrical framework we model the location $\mu_c$ of the circular
variable and the mean $\mu_l$ or scale $\nu$ of the linear variable. Note that
to prevent confusion we distinguish between the circular location and linear
mean using subscripts. In this paper we use covariates to predict the circular
and linear component using the following general type of prediction equations
for a cylindrical model with $q$ covariates for the linear component and $p$
covariates for the circular component:
\begin{equation}\label{circpredgen}
\hat{\theta_i} = g(\beta_0 + \beta_1z_1 + \dots + \beta_qz_p),
\begin{equation}\label{linpredgen}
\hat{y_i} = h(\gamma_0 + \gamma_1x_1 + \dots + \gamma_qx_q),
where $g()$ and $h()$ are link functions, $\boldsymbol{\beta} = (\beta_0, \dots,
\beta_p)$ is a vector of regression coefficients for the circular variable and
$\boldsymbol{\gamma} = (\gamma_0, \dots, \gamma_q)$ is a vector of regression
coefficients for the linear variable. The link functions that are chosen depend
on the type of distribution we choose for the linear and circular component. In
this paper we use either an identity link or an exponential function (when
modelling a scale parameter) for the linear component. For the circular
component the link functions are specific to circular data and include
arctangent functions. The specific type of link function and prediction
equations used will be introduced in the respective descriptions of the models
used in this paper.\newline
\indent In addition to modelling the circular and linear component of a
cylindrical variable separately using covariates we also model the relation
between them. In the literature, several methods have been put forward to do so.
Some of these are based on regressing the linear component onto the circular
component by including $\sin(\theta_i)$ and $\cos(\theta_i)$ into the prediction
equation in \eqref{linpredgen} [@mardia1978model;@johnson1978some;
@mastrantonio2015bayesian]. The linear component is thus modeled using the sine
and cosine of the circular component. Others model the relation in a different
way, e.g. by specifying a multivariate model for the linear and circular
variable and modelling their covariance matrix [@mastrantonio2018joint] or by
proposing a joint cylindrical distribution. For example, @abe2017tractable
introduce a cylindrical distribution based on a Weibull distribution for the
linear component and a sine-skewed von Mises distribution for the circular
component and link these through their respective shape and concentration
parameters.

@@

\subsection{Four Cylindrical Regression Models}\label{Models}
One of the goals of this paper is to show the benefits of cylindrical methods
for the analysis of circumplex data and cylindrical data in psychology in
general. To do so we focus on four cylindrical models. The models were selected
for their relatively low complexity and the ease with which a regression
structure could be incorporated. But also because they show different ways of
modelling the linear and circular component and thereby illustrate a wider range
of cylindrical models available in the literature. As outlined in the previous
section, the cylindrical models contain a set of $q$ predictors $\boldsymbol{x}
= x_1, \dots, x_q$ and $p$ predictors $\boldsymbol{z} = z_1, \dots, z_p$ for the
linear and circular components, \(Y\) and \(\Theta\), respectively. The first
two models are based on a construction by @mastrantonio2015bayesian, while the
other models are extensions of the models from @abe2017tractable and
@mastrantonio2018joint. The four cylindrical  models are introduced separately
in the subsections below. However, to provide a more succinct overview and
comparison of the four models, Table \ref{TableModels} gives an overview of the
similarities and differences between the models.

\begin{sidewaystable}[h]
\centering
\caption{Comparison of the four cylindrical regression models} 
\begin{tabular}{lllll}
  \noalign{\smallskip}\hline\noalign{\smallskip}
\multicolumn{1}{l}{Aspect} & CL-PN & CL-GPN  & Abe-Ley  & GPN-SSN \\ \hline\noalign{\smallskip}
$\Theta$ & &&&\\
$\:\:$Distribution& PN & GPN & Sine-skewed vM & GPN\\
$\:\:$Domain & $[0, 2\pi)$ & $[0, 2\pi)$ & $[0, 2\pi)$ & $[0, 2\pi)$\\
$\:\:$Shape & symmetric, & asymmetric, & asymmetric, & asymmetric, \\
            & unimodal  & multimodal & unimodal   & multimodal \\\hline\noalign{\smallskip}
$Y$& &&&\\
$\:\:$Distribution & Normal & Normal & Weibull & skewed-Normal\\
$\:\:$Domain & $(-\infty, + \infty)$ & $(-\infty, + \infty)$ & $(0, + \infty)$ & $(-\infty, + \infty)$\\
$\:\:$Shape & symmetric, & symmetric, & asymmetric, & asymmetric, \\
            & unimodal  & unimodal  & unimodal   & unimodal\\\hline\noalign{\smallskip}
$\Theta$-$Y$ dependence &                                   &                                   & & \\
                        & $Y$ regressed on                  & $Y$ regressed on                  & circular concentration $\kappa$  & covariance \\
                        & $\sin(\Theta)$ and $\cos(\Theta)$ & $\sin(\Theta)$ and $\cos(\Theta)$ & and linear scale $\alpha$ & matrix\\\hline
\multicolumn{5}{l}{Note: PN and GPN refer to the projected normal and general projected normal distribution.}\\
\multicolumn{5}{l}{vM refers to the von-Mises distribution}\\

\hfil [Table \ref{TableModels} about here] \hfil

\subsubsection{The Modified Circular-Linear Projected Normal (CL-PN) and Modified Circular-Linear General Projected Normal (CL-GPN) Models}\label{CL-(G)PN}
Following @mastrantonio2015bayesian we consider two models where the prediction
equation for the linear component is specified by combining \eqref{linpredgen},
with an identity link function and including $\sin(\theta_i)$ and
$\cos(\theta_i)$ as follows:
\begin{equation}\label{linpredCLPNCLGPN}
\hat{y_i} = \gamma_0 + \gamma_{cos}*\cos(\theta_i)*r_i + \gamma_{sin}*\sin(\theta_i)*r_i + \gamma_1*x_1 + \dots + \gamma_q*x_q,
\end{equation}
\noindent where $r_i$ is a realization of the unobserved random variable
\(R\geq0\) that will be introduced below, \(\gamma_0, \gamma_{cos},
\gamma_{sin}, \gamma_1, \dots, \gamma_q\) are the intercept and regression
coefficients and \(x_1, \dots, x_q\) are the \(q\) covariates for the prediction
of the linear component. We assume a normal $N(\mu_l, \sigma^2)$ distribution
for the linear component.\newline
\indent For the circular component we assume either a projected normal (PN) or a
general projected normal (GPN) distribution. These distributions arise from a
projection of a distribution defined in bivariate space onto the circle. Figure
\ref{projection} represents this projection. In the left plot of Figure
\ref{projection} we see realizations $(s_i^{I}, s_i^{II})$ from the bivariate
normal variable \(\boldsymbol{S}\) that in the middle plot are projected to form
the circular component \(\Theta\) in the right plot. Mathematically the relation
between \(\boldsymbol{S}\) and \(\Theta\) is defined as follows:
\begin{equation}\label{projection_eq}
\boldsymbol{S} = \begin{bmatrix} S^{I} \\ S^{II} \end{bmatrix} = \begin{bmatrix} R \cos (\Theta) \\  R\sin (\Theta) \end{bmatrix},
\end{equation}
\noindent where \(R = \mid\mid \boldsymbol{S} \mid\mid\), the Euclidean norm of
\(\boldsymbol{S}\), that is represented by the lines connecting the bivariate
datapoints to the origin in the middle plot. We call $\boldsymbol{S}$ the
augmented representation of the circular component. It is a variable that in
contrast to \(\Theta\) is not observed and thus considered latent or auxiliary.
This then means that we do not model \(\Theta\) directly but indirectly through
$\boldsymbol{S}$.

\hfil [Figure \ref{projection} about here] \hfil

\indent For both the PN and GPN distributions the circular location parameter
\(\mu_c\in [0, 2\pi)\) is modeled.\footnote{Note that for the CL-GPN
model the circular location parameter also depends on the variance-covariance
matrix and the circular predicted values should be computed using numerical
integration or Monte Carlo methods because a closed form expression for the mean
direction is not available.} The prediction equation for the circular component is specified by using a double arctangent link function in \eqref{circpredgen} as follows:
\begin{equation}\label{circpredCLPNCLGPN}
\hat{\theta_i} = \text{atan2}(\boldsymbol{\beta^{II}}\boldsymbol{z}_i, \boldsymbol{\beta^{I}}\boldsymbol{z}_i)
\end{equation}
\noindent where $\boldsymbol{\beta}^{I} = (\beta_0^{I},
\beta_1^{I}, \dots, \beta_p^{I})$ and
$\beta^{II} = (\beta_0^{II},
\beta_1^{II}, \dots, \beta_p^{II})$ are vectors with
intercepts and regression coefficients for the prediction of $S^{I}$ and
$S^{II}$ and $\boldsymbol{z}_i$ is a vector with predictor values for each
individual $i \in 1, \dots, n$ where $n$ is the sample size. Note that as a
result of the augmented representation of the circular component we have two sets
of regression coefficients and intercepts, in contrast to a single set in
\eqref{circpredgen}. This leads to problems when we want to interpret the effect
of a covariate on the circle. A circular regression line for simulated data is shown in Figure
\ref{circregline}, with covariate values on the x-axis and the predicted
circular component on the y-axis. As can be seen it is of a non-linear character
meaning that the effect of a covariate is different at different values of the
covariate. A circular regression line is usually described by the slope at the
inflection point, the point at which the slope of the regression line starts
flattening off (indicated with a square in Figure \ref{circregline}). By
default, the parameters from the PN and GPN models do not directly describe this
inflection point. For the PN distribution however, @CremersMulderKlugkist2017
solved this interpretation problem. They introduce a new parameter $b_c$ that describes the slope at
the inflection point of the regression line. For the GPN distribution the
interpretation problem however remains.

\hfil [Figure \ref{circregline} about here] \hfil

\indent The main difference between the PN and GPN distribution lies in the
definition of their covariance matrix. For the PN distribution this is an
identity matrix, causing the distribution to be unimodal and symmetric, whereas
for the GPN distribution \(\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \xi^2
& \xi\\ \xi & 1 \end{bmatrix}\) where \(\xi,\tau \in (-\infty, +\infty)\),
allowing for multimodality and asymmetry/skewness as illustrated in Figure \ref{plotGPN}.

\hfil [Figure \ref{plotGPN} about here] \hfil

\indent Both the CL-PN and CL-GPN models are estimated using Markov Chain Monte
Carlo (MCMC) methods based on @nunez2011bayesian, @wang2012directional and
@hernandez2016general for the regression of the circular component. A detailed
description of the Bayesian estimation and MCMC samplers can be found in the
Supplementary Material.

\begin{figure}
\centering
\includegraphics[width = \textwidth]{Plots/plotprojecting.pdf}
\caption{Plot showing the projection of datapoints in bivariate space, $\boldsymbol{S}$, (left) to the circle (right). The lines connecting the bivariate datapoints to the circular datapoints represent the euclidean norms of the bivariate datapoints, realizations of the random variable $R$.}
\label{projection}
\end{figure}

CompI  <- 1.1 + 2.5*Xreal + rnorm(100, 0.1,1)
CompII <- 2.1 + -0.4*Xreal + rnorm(100, 0.1,1)

predCompI  <- 1.1 + 2.5*X
predCompII <- 2.1 + -0.4*X

axval <- ((1.1*2.5)+(2.1*-0.4))/(1.1^2 + 2.1^2)
acval <- atan2(2.1 + -0.4*axval, 1.1 + 2.5*axval)

\begin{figure}[]
  \includegraphics[width = \textwidth]{Plots/circregline.pdf}
  \caption{Circular regression line for the relation between a covariate and a circular component with the simulated data the regression line was fit to. The square indicates the inflection point of the regression line.}
  \label{circregline}
\end{figure}

```{r plotGPN, cache = TRUE, warning = FALSE, dev = "tikz",fig.ext="svg", echo = FALSE, results = FALSE}

require(MASS)
require(circular)

y1 = mvrnorm(1000, c(2,2), matrix(c(1^2 + 0^2, 0, 0, 1), nrow = 2, ncol = 2))
y2 = mvrnorm(1000, c(2,2), matrix(c(5^2 + 0.9^2, 0.9, 0.9, 1), nrow = 2, ncol = 2))
y3 = mvrnorm(1000, c(2,2), matrix(c(10^2 + -0.5^2, -0.5, -0.5, 1), nrow = 2, ncol = 2))

theta_1 = atan2(y1[,2], y1[,1])
theta_2 = atan2(y2[,2], y2[,1])
theta_3 = atan2(y3[,2], y3[,1])

tikz("Plots/plotGPN.tex", standAlone =TRUE, height = 3, width = 9, pointsize = 12, engine = "pdftex")

par(mfrow=c(1,3),
    oma=c(0,0,0,0),
    mar=c(0,0,0,0))

plot(as.circular(theta_1), stack = TRUE,
     main = " ", ylab = " ", xlab = " ", shrink = 1.3, axes = FALSE)
axis.circular(at = 0, labels = "0")
plot(as.circular(theta_2), stack = TRUE,
     main = " ", ylab = " ", xlab = " ", shrink = 1.3, axes = FALSE)
axis.circular(at = 0, labels = "0")
plot(as.circular(theta_3), stack = TRUE,
     main = " ", ylab = " ", xlab = " ", shrink = 1.3, axes = FALSE)
axis.circular(at = 0, labels = "0")

#Close the device
dev.off()

# Compile the tex file
tools::texi2dvi("Plots/plotGPN.tex",pdf=TRUE)

\begin{figure}
\centering
\includegraphics[width = \textwidth]{Plots/plotGPN.pdf}
\caption{Plot showing 1000 samples from three GPN distributions with mean vector $(2,2)$ and $\tau = 1,5,10$ and $\xi = 0, 0.9, -0.5$ from left to right}
\label{plotGPN}
\end{figure}

\subsubsection{The Modified Abe-Ley Model}\label{WeiSSVM}
This model is an extension of the cylindrical model introduced in
@abe2017tractable to the regression context. The circular dependence between the
  linear and circular component, $Y$ and $\Theta$, is defined through a joint
  density function as follows:

\begin{equation}\label{WeiSSVMdensity}
f(\theta, y) = \frac{\alpha\nu^\alpha}{2\pi\cosh(\kappa)}
                 (1 +\lambda\sin(\theta - \mu_c))
                 y^{\alpha-1}
                 \exp[-(\nu y)^{\alpha}(1-\tanh(\kappa)\cos(\theta - \mu_c))],
\end{equation}

\noindent where \(\alpha > 0\) is a linear shape parameter, \(\kappa > 0\) and
\(\lambda \in [-1, 1]\) are circular concentration and skewness parameters
respectively. The Abe-Ley density thus concerns a combination of a Weibull
distribution, with scale parameter \(\nu>0\) and shape parameter \(\alpha\), for
the linear component and a sine-skewed von Mises distribution, with location
parameter \(\mu_c\in [0, 2\pi)\), concentration parameter \(\kappa>0\) and
skewness \(\lambda \in [-1,1]\), for the circular component. In contrast to the
CL-PN and CL-GPN models, the linear component \(Y\) is in this model defined
only on the positive real half-line \([0, + \infty)\) and thus can not be
negative. Our modification of the Abe-Ley model occurs at the level of the
linear scale parameter \(\nu>0\) and circular location parameter \(\mu_c\in [0,
2\pi)\), both of which we will model using covariates.\newline
\indent The prediction equation for the circular component equals:
\begin{equation}\label{circpredcAL}
\hat{\theta}_i = \beta_0 + 2\tan^{-1}(\boldsymbol{\beta}\boldsymbol{z}_i)
\end{equation}
Note that we have taken out the intercept $\beta_0$ from the link function $g() =
2\tan^{-1}$ in \eqref{circpredgen}. The vector $\boldsymbol{\beta}$ thus does
not contain an intercept. We do not directly predict the linear component. The
conditional distribution for the linear component is Weibull, meaning that we can
use methods from survival analsis to interpret the effect of a predictor. In
survival analysis a 'survival' function is used in which time is plotted against
the probability of survival of subjects suffering from a specific medical
condition. This probability is computed using the 'survival-function' defined as
\begin{equation}\label{survivalfunc}
\exp(-\alpha
y_i^{\hat{\nu}_i(1-\tanh(\kappa)\cos(\theta_i - \hat{\theta}_i))^{1/\alpha}}),
\end{equation}
\noindent with \(\hat{\nu}_i = \exp(\boldsymbol{\gamma}\boldsymbol{x}_i)\). Note that $\boldsymbol{x}_i$ also includes a 1 to be able to estimate the intercept $\gamma_0$. From
the survival function we also see that the circular concentration parameter
\(\kappa\) and the linear shape parameter \(\alpha\) regulate the
circular-linear dependence in the Abe-Ley model. In the exponent or power of the linear outcome, the prediction error on the circle, $\cos(\theta_i - \hat{\theta_i})$, is multiplied by $\tanh{\kappa}$, and raised to the power $1-\alpha$.\newline
\indent We can use numerical optimization (Nelder-Mead) to find solutions for
the maximum likelihood (ML) estimates for the parameters of the model.

\subsubsection{Modified Joint Projected and Skew Normal Model (GPN-SSN)}\label{CL-GPN_multivariate}
This model is an extension of the cylindrical model introduced by
@mastrantonio2018joint to the regression context. Although the model may contain
several circular and linear components we will restrict ourselves to one
circular and one linear component in this paper. The circular component is
modelled by a GPN distribution, as in the CL-GPN model, while the linear component is modelled by a skew
normal distribution [@sahu2003new]. Because the GPN distribution is modelled
using a so-called augmented representation (see \eqref{projection_eq}) it is
convenient to use a similar tactic for modelling the skew normal distribution.
As in @mastrantonio2018joint, dependence between the linear and circular
component is created by modelling the augmented representation of \(\Theta\) and
\(Y\) together in a 3 dimensional normal distribution. The joint density of the model is then represented by:
\begin{equation}\label{YDThetarjoint} 
f(\theta, r, y, d) = 2\phi_{3}((\boldsymbol{s}^t,
y^t)^t \mid \boldsymbol{M} + (0,0,\lambda d)^t, \boldsymbol{\Sigma})
\phi(d \mid 0, 1)r, 
\noindent where $\phi$ represents the normal probability density function, \(\boldsymbol{s} = (r(\cos(\theta), \sin(\theta)))^t\), $d$ is
the augmented representation of the linear component, $\boldsymbol{M} =
\boldsymbol{B}^t\boldsymbol{X}$, $\boldsymbol{B}_{(g + 1) \times 3}$ is
defined below, $\boldsymbol{X}$ is a design matrix and $\boldsymbol{\Sigma} =
\left ( \begin{matrix} \boldsymbol{\Sigma}_s & \boldsymbol{\Sigma}_{sy} \\
\boldsymbol{\Sigma}_{sy}^t & \boldsymbol{\Sigma}_y \\ \end{matrix} \right )$ a
variance-covariance matrix. The matrices \(\boldsymbol{\Sigma}_s\) and \(\boldsymbol{\Sigma}_y\) are the
covariance matrices for the variances of and covariances between the augmented
representations of the circular and linear component respectively. Note that because we only have one circular and one linear component in this paper, \(\boldsymbol{\Sigma}_s\) is a two by two matrix and \(\boldsymbol{\Sigma}_y\) is a scalar representing the variance of the augmented linear component. The matrix
\(\boldsymbol{\Sigma}_{sy}\) contains covariances between the augmented
representations of the circular and linear component. The matrix with
regression coefficients and intercepts, $\boldsymbol{B}$ is defined as follows:
\begin{equation}\label{regmatGPNSSN}
\boldsymbol{B} = \begin{bmatrix}
\beta_{0_{s^{I}}} & \beta_{0_{s^{II}}} & \beta_{0_{y}}\\
\beta_{1_{s^{I}}} & \beta_{1_{s^{II}}} & \beta_{1_{y}}\\
\vdots & \vdots & \vdots \\
\beta_{g_{s^{I}}} & \beta_{g_{s^{II}}} & \beta_{g_{y}} \end{bmatrix}.
\noindent Note that in contrast to the previous three models where we use $p$
and $q$ covariates for the circular and linear component respectively, we use $g
= p = q$ covariates for both components in the GPN-SSN model. Unlike the other
models, where $\mu_l$ and $\mu_c$ are modelled separately, the GPN-SSN thus has
a shared mean vector $\boldsymbol{M}$ and variance-covariance matrix
$\boldsymbol{\Sigma}$ for the linear and circular component, much like having
multiple outcomes in a MANOVA (multivariate analysis of variance) model. This
implies that the covariances in the variance-covariance matrix describe the
dependence between the circular and linear component.\newline
\indent The prediction equation for the circular component\footnote{Note that for the  GPN-SSN model the predicted circular component also depends on the variance-covariance matrix and the circular predicted values should be computed using numerical integration or Monte Carlo methods because a closed form expression for the mean direction is not available.} is similar to \eqref{circpredCLPNCLGPN} and equal to:
\begin{equation}\label{circpredGPNSSN}
\hat{\theta}_i = \text{atan2}(\boldsymbol{\beta^{II}}\boldsymbol{z}_i, \boldsymbol{\beta^{I}}\boldsymbol{z}_i),
\noindent where $\boldsymbol{\beta^I} = (\beta_{0_{s^{I}}}, \beta_{1_{s^{I}}}, \dots, \beta_{g_{s^{I}}})$ and $\boldsymbol{\beta^{II}} = (\beta_{0_{s^{II}}}, \beta_{1_{s^{II}}}, \dots, \beta_{g_{s^{II}}})$.
For the linear component we subsititute $h()$ in \eqref{linpredgen} for an identity link and $\boldsymbol{\gamma}$ for $\boldsymbol{\beta}$:
\begin{equation}\label{linpredGPNSSN}
\hat{y}_i = \boldsymbol{\beta}\boldsymbol{x}_i,
\noindent where $\boldsymbol{x}_i = \boldsymbol{z}_i$ and $\boldsymbol{\beta} = (\beta_{0_{y}}, \beta_{1_{y}}, \dots, \beta_{g_{y}})$.\newline
\indent We estimate the model using MCMC methods. A detailed description of these
methods is given in the Supplementary Material.
